Place of death in NRS COVID-19 deaths

### 2021-02-09 covid_nrs_place_of_death.Rmd
added HB and LA data using 2week moving average; started an overall yearly look at data but need to include yearly range from historical data, plus a split of coivd-as underlying cause and one of factors in death

# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
#   geom_polygon() +
#   coord_map()

```

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
# figure_deaths_at_home <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

@@

merged_deaths_la %>%
  filter(place_of_death == "Home & other non-institution") %>%
  filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

@@

>>>>

# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
#   geom_polygon() +
#   coord_map()
```

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

## Deaths at home with 2-week moving average, by health board

```{r, fig.width=12, fig.height=9, warning=FALSE}
moving_average_text <- "All figures represent deaths averaged over past 2 weeks; this includes historical range and historical mean."

figure_deaths_at_home_2week_average_by_health_board <-
  merged_deaths_hb %>%
  filter(place_of_death == "Home & other non-institution") %>%
  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y") +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_health_board",.x), plot = figure_deaths_at_home_2week_average_by_health_board, width = 12, height = 9, dpi = 300)
)

figure_deaths_at_home_2week_average_by_health_board
```

@@

# merged_deaths_la %>%
#   filter(place_of_death == "Home & other non-institution") %>%
#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
#                names_to = "deaths_type", values_to = "deaths") %>% 
#   ggplot(aes(x = week_number_run_over)) +
#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
#   geom_line(aes(y = deaths, colour = deaths_type)) +
#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
#   geom_vline(xintercept = 54, linetype = "dotted") +
#   scale_x_continuous(breaks = seq(0,60,by=10)) +
#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
#   theme_minimal() +
#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
#                       labels = c("All deaths", 
#                                  "Non Covid-19 deaths"),
#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
#                       ) +
#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
#   labs(x = "Week number",
#        y = "Deaths (N)",
#        subtitle = date_range_text,
#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
#        colour = "Cause of death"
#        )

@@

## Deaths at home with 2-week moving average, by local authority

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_2week_average_by_local_authority <-
  merged_deaths_la %>%
  filter(place_of_death == "Home & other non-institution") %>%
  pivot_longer(cols = c(ma2w_deaths_all_causes, ma2w_deaths_non_covid), 
               names_to = "deaths_type", values_to = "ma2w_deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = ma2w_deaths_min, ymax = ma2w_deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = ma2w_deaths, colour = deaths_type)) +
  geom_line(aes(y = ma2w_deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~ref_area, scales = "free_y") +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(moving_average_text, vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )
# 
# merged_deaths_la %>%
#   filter(place_of_death == "Home & other non-institution") %>%
#   filter(ref_area %in% c("Moray","Highland", "Fife", "Glasgow City")) %>%
#   pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
#                names_to = "deaths_type", values_to = "deaths") %>% 
#   ggplot(aes(x = week_number_run_over)) +
#   geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
#   geom_line(aes(y = deaths, colour = deaths_type)) +
#   geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
#   geom_vline(xintercept = 54, linetype = "dotted") +
#   scale_x_continuous(breaks = seq(0,60,by=10)) +
#   facet_wrap(~ref_area, scales = "free_y", nrow = 2) +
#   theme_minimal() +
#   scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
#                       labels = c("All deaths", 
#                                  "Non Covid-19 deaths"),
#                       guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
#                       ) +
#   theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
#   labs(x = "Week number",
#        y = "Deaths (N)",
#        subtitle = date_range_text,
#        caption = paste0(c(vertical_line_text, vertical_scale_differs_between_panels_text, source_nrs_text), collapse="\n"),
#        colour = "Cause of death"
#        )

walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0("./outputs/figure_deaths_at_home_2week_average_by_local_authority",.x), plot = figure_deaths_at_home_2week_average_by_local_authority, width = 12, height = 9, dpi = 300)
)

figure_deaths_at_home_2week_average_by_local_authority
```

@@

## Overall deaths at home 2020 by cause of death

```{r}
# TODO: add yearly range for all deaths
# TODO: split data by covid-underlying cause of desath & covid-factor in death
merged_data_overall %>%
  filter(sex=="all", age=="all") %>%
  filter(year==2020) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  group_by(ref_area, year) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
  )
```
